---
title: "Pre-Lab: Installing pavo and Friends"
subtitle: "Complete BEFORE Wednesday May 6 — Lab 5: Pavo Introduction"
author: "EEB 187 — Ecology & Evolution of Color"
format:
  html:
    toc: true
    code-overflow: wrap
  pdf:
    toc: true
---

## Why this matters

Wednesday's lab is a hands-on R session. **You must arrive with a working R + RStudio + pavo installation.** If you try to install during class, you will fall behind and miss the live walkthrough. Allocate about **30 minutes** the day before lab to do this. Most of that time is downloads.

If you hit problems, post in the EEB 187 channel — there's almost certainly a classmate fighting the same error.

---

## What you need to install

| Tool | Version | Why |
|---|---|---|
| **R** | ≥ 4.3 | The language pavo runs on |
| **RStudio** | Recent (2024+) | The editor we'll use |
| **pavo** | ≥ 2.9 | Color pattern analysis |
| **tidyverse** | Recent | Data wrangling + ggplot |
| **magick** | Recent | Image I/O backend pavo uses |

---

## Step 1: Install R (if you don't have it)

- **macOS:** [https://cran.r-project.org/bin/macosx/](https://cran.r-project.org/bin/macosx/) — get the `.pkg` for your chip (Apple Silicon = "arm64", Intel = "x86_64")
- **Windows:** [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/) — get the `.exe`
- **Linux:** use your distro's package manager (`apt install r-base`, etc.)

Verify by opening a terminal and running:

```bash
R --version
```

You should see something like `R version 4.3.x` or higher.

---

## Step 2: Install RStudio

[https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/) — pick the free desktop version. Install. Open it once to confirm it launches.

---

## Step 3: Install the R packages

Open RStudio. In the Console pane (bottom-left), paste and run:

```{r}
#| eval: false
install.packages(c("pavo", "tidyverse", "magick", "knitr"))
```

This will take **5–15 minutes** depending on your connection. You'll see a lot of compilation output. That's normal. Wait for the prompt (`>`) to come back.

::: {.callout-warning}
**On macOS:** if you see errors about `Cellar`, `gfortran`, or `clang`, you need Apple's command-line tools first. Run in Terminal: `xcode-select --install`, then re-run the R install.

**On Windows:** if pavo fails to install, you may need to install Rtools first: [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/)
:::

---

## Step 4: Smoke test

In the RStudio Console, paste and run **all** of this:

```{r}
#| eval: false
library(pavo)
library(magick)
library(tidyverse)

# Load a test image bundled with pavo
data(sicalis)

# Run a quick check of the pavo version
packageVersion("pavo")

cat("\nIf you see a version number above (≥ 2.9.0) and no errors,\n")
cat("you are ready for Wednesday's lab.\n")
```

**You should see:**

```
[1] '2.9.0'   # or higher

If you see a version number above (≥ 2.9.0) and no errors,
you are ready for Wednesday's lab.
```

If you got that, **stop**. You're done.

---

## Step 5 (optional but recommended): Create a lab project folder

Pick a place on your computer where you keep coursework — Documents, Desktop, wherever. Make a folder called `EEB187-pavo-lab`. On Wednesday I'll have you download a ZIP from BruinLearn that you'll unpack into this folder. Doing this in advance saves 90 seconds of fumbling on the day.

```
Documents/
  EEB187-pavo-lab/      ← create this empty folder now
```

---

## Step 6 (RECOMMENDED): Install the segmentation tools

Your fish images need their backgrounds removed before pavo can analyze them cleanly. The lab uses two command-line tools, both invoked from R via a one-line helper. **Install both before lab if you can** — installing during lab will eat your work time.

### macOS

::: {.callout-note title="Don't have Homebrew yet? Most students won't."}
Homebrew (`brew`) is the macOS package manager — a one-time install that makes installing things like ImageMagick a one-liner forever after. **You only need to do this once on your laptop.**

**Install Homebrew (5 minutes):**

1. Open **Terminal** (`Applications → Utilities → Terminal`, or hit ⌘-Space and type "Terminal").
2. Paste this command and press return:

   ```bash
   /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
   ```

3. It will ask for your **macOS password** (the one you use to log into your Mac). Type it — characters won't show as you type, that's normal — and press return.
4. The installer downloads and configures Homebrew. Watch for a message at the very end labeled **"Next steps"**. On Apple-Silicon Macs (M1/M2/M3/M4) it will tell you to run two more commands that look like this:

   ```bash
   echo 'eval "$(/opt/homebrew/bin/brew shellenv)"' >> ~/.zprofile
   eval "$(/opt/homebrew/bin/brew shellenv)"
   ```

   **Run those exact commands** (copy from your Terminal output, not from this guide — your username will be different). They put `brew` on your `PATH` so future Terminal sessions can find it.

5. Confirm it works:

   ```bash
   brew --version           # should print "Homebrew 4.x.x"
   ```

If `brew --version` says "command not found," your shell hasn't been told about Homebrew yet. **Quit Terminal and reopen it**, then try again. If still no luck, go back and re-do step 4 (the `eval` line).

Full official instructions, in case anything looks different on your machine: <https://brew.sh>
:::

Once Homebrew is in place:

```bash
# In Terminal:
brew install imagemagick
pip install "rembg[cpu,cli]"              # or pipx install "rembg[cpu,cli]"
```

Verify both work:

```bash
magick -version          # should print "Version: ImageMagick 7.x..."
rembg --help             # should print rembg's usage text
```

### Windows

Install [ImageMagick from the official site](https://imagemagick.org/script/download.php#windows) (the "Windows Binary Release" installer), then in Command Prompt:

```bash
pip install "rembg[cpu,cli]"
```

### Linux

```bash
sudo apt install imagemagick
pip install "rembg[cpu,cli]"
```

### Recommended alternative: install `uv` instead of pip-installing rembg

Several students will have trouble with `pip install "rembg[cpu,cli]"` because rembg's deep-learning dependencies (onnxruntime, opencv) don't always have prebuilt wheels for the latest Python. There's a much smoother path that **handles all of this automatically**:

```bash
# macOS:
brew install uv

# Linux:
curl -LsSf https://astral.sh/uv/install.sh | sh

# Windows:
powershell -ExecutionPolicy ByPass -c "irm https://astral.sh/uv/install.ps1 | iex"
```

That's it. With `uv` installed, the `segment_fish()` helper in lab will run rembg via `uvx --from "rembg[cpu,cli]" rembg ...` — uv downloads a compatible Python and rembg automatically the first time you call the helper, and caches both for future calls. **No more pip install errors, no Python version mismatches.**

`segment_fish()` tries `uvx` first and falls back to a direct `rembg` binary, so if you got rembg installed via pip and it's on PATH, that works too.

::: {.callout-warning}
## Common roadblock 1 — Windows + pip install fails on rembg

If you tried `pip install "rembg[cpu,cli]"` on Windows and saw errors about Visual C++ build tools, ONNX runtime, or opencv: **don't fight it**. Install `uv` instead (one PowerShell command above) and skip pip entirely. `segment_fish()` will use `uvx` automatically.
:::

::: {.callout-warning}
## Common roadblock 2 — "rembg installed but R can't find it"

If `rembg --help` works in your terminal but R says `rembg failed (or not installed)`: R doesn't see your full PATH. Check what R sees:

```r
Sys.getenv("PATH")
```

If `/opt/homebrew/bin` (macOS Homebrew) or `~/.local/bin` (pip user installs) isn't in there, **install `uv` and use that path instead** — `uvx` is more discoverable. Or set the PATH in `~/.Renviron`:

```
PATH=/opt/homebrew/bin:/usr/local/bin:/usr/bin:/bin:/usr/sbin:/sbin:~/.local/bin
```

Restart RStudio for the change to take effect.
:::

::: {.callout-tip}
## Common roadblock 3 — "I had to fall back to manual cropping; how does that affect my answers?"

If neither `uv` nor `rembg` work for you, the worksheet's **manual-crop path** (Step 1.5 Path 2) is fully supported. Your A values will be less dramatic than the demo (closer to 1 across the board) because some background pixels will pollute the analysis. **The directional A claim still holds** — vertical-bar fish > 1, horizontal-stripe fish < 1 — just with smaller effect sizes. Note in your worksheet which path you used; the rubric accommodates both.
:::

### If install fails entirely

Come to lab anyway with your three lineage fish images saved locally. Use the manual-crop path. The lab is still teachable; just plan for the smaller-magnitude metrics described above.

---

## Step 7 (REQUIRED): Identify three fish from your assigned lineage

Your group has been assigned a reef fish lineage (typically a family). Before lab on Wednesday, **pick three species from your lineage that show three visually different pattern types** — for example:

- One with strong **vertical bars** (you'll predict A > 1)
- One with a **horizontal stripe or band** (you'll predict A < 1)
- One that's **spotted, reticulate, or solid** (you'll predict A ≈ 1)

If your lineage doesn't span all three pattern types, pick the three with the most different patterns you can find in it.

**Image source — use whichever your group has been working with:**

- **FishView server** — if your group has been curating images there, use that
- **FishBase** ([fishbase.se](https://fishbase.se)) or **Wikimedia Commons** — search by species name; download a clear, lateral-view photo
- **Fallback:** the lab bundle includes pre-curated images at `images/student-fish/<your-lineage>/` — five species per lineage, ready to use

Save your three images locally before lab — either in your `EEB187-pavo-lab` folder or somewhere you can find them quickly. **Don't show up empty-handed.** If you don't know your lineage assignment, ask in the channel by Tuesday evening.

---

## Common errors and fixes

**`there is no package called 'pavo'`** — `install.packages("pavo")` didn't actually run. Re-run it; check for a typo.

**`unable to load shared object … magick.so`** — On macOS, install ImageMagick via Homebrew first: `brew install imagemagick`, then reinstall the R `magick` package. On Linux: `sudo apt install libmagick++-dev`.

**`installation of package 'XXX' had non-zero exit status`** — The package failed to compile. Scroll up in the install output to see the actual error. Most common: missing system dependency. Google the error or post in the channel.

**RStudio crashes on startup** — Reinstall RStudio. If still failing, your R install is broken; reinstall R, then RStudio.

---

## What we'll do in lab

You don't need to read up on pavo or do prep beyond Steps 1–7 above. Wednesday's lab walks you through everything from scratch:

1. **Demo block** — three reef fish from three families: sergeant major (vertical bars), cleaner wrasse (horizontal stripe), mandarin fish (reticulate). All three are pre-segmented for you.
2. **Predict** which pattern statistic will tell them apart (spoiler: the **A** statistic)
3. Run pavo's `classify()` and `adjacent()` on all three demos; compare predictions to measurements
4. **Your turn** — segment your three lineage fish using `segment_fish()` (or manual crop), then run the same pipeline
5. Compare your three fish's metrics to the demo fish — does the A statistic correctly classify your fish by pattern orientation, the way it did for the demo?
6. Submit a worksheet with three metric tables, a comparison answer, and a biological interpretation

---

## Need help?

Post in the EEB 187 channel before Wednesday. Don't show up Wednesday with a broken install — you'll spend the lab on installation instead of biology.
